Automatic prostate cancer detection through \acs*{dce}-\acs*{mri} images: all you need is a good normalization

### 2016-08-29 01_introduction/introduction.tex
Add fab correction

\emph{Semi-quantitative} approaches are rather mathematical than pharmacokinetic modelling since that no pharmacokinetic assumption regarding the relation between the \ac{mri} signal and the contrast agent~\citep{huisman2001accurate,gliozzi2011phenomenological}.
These methods offer the advantages to not require any knowledge about the \ac{mri} sequence nor any conversion from signal intensity to concentration.
However, the heuristic approach proposed by \citeauthor{huisman2001accurate} requires an initial estimate of the noise standard deviation of the signal as well as manual tuning.

@@

In this work, we propose a fully automatic normalization method for \ac{dce}-\ac{mri} that reduces the inter-patient variability of the data.
The benefit and simplicity of our approach will be shown by classifying the whole normalized \ac{dce}-\ac{mri} signal and comparing with the state-of-the-art quantitative and semi-quantitative methods.
Additionally, we will show that using this normalization approach in conjunction with the quantitative methods improves the classification performance in for most of the models.

>>>>

\emph{Semi-quantitative} approaches are rather mathematical than pharmacokinetic modelling since that no pharmacokinetic assumption regarding the relation between the \ac{mri} signal and the contrast agent are made~\citep{huisman2001accurate,gliozzi2011phenomenological}.
These methods offer the advantages to not require any knowledge about the \ac{mri} sequence nor any conversion from signal intensity to concentration.
However, the heuristic approach proposed by \citeauthor{huisman2001accurate} requires an initial estimate of the noise standard deviation of the signal as well as some manual tuning.

@@

In this work, we propose a fully automatic normalization method for \ac{dce}-\ac{mri} that reduces the inter-patient variability of the data.
The benefit and simplicity of our approach will be shown by classifying the whole normalized \ac{dce}-\ac{mri} signal and comparing with the state-of-the-art quantitative and semi-quantitative methods.
Additionally, we will show that using this normalization approach in conjunction with the quantitative methods improves the classification performance for most of the models.

### 2016-08-29 02_methods/methods.tex
Add fab correction

In \ac{dce}-\ac{mri}, the intensity \ac{pdf} of prostate gland does not follow a unique type of distribution such as Rician or Gaussian distribution, as shown in Fig.\,\ref{subfig:pathhist}.
Indeed, the inter-patient variations are more complex due to the temporal acquisition.
A better representation to observe these variations is to represent the intensity \ac{pdf} of the \ac{dce}-\ac{mri} data over time using a heatmap representation as shown in Fig.\,\ref{subfig:pathhist}.
Analyzing this heatmap representation across patients (see Fig.\,\ref{subfig:pat2}), the following variations are highlighted:
(i) an intensity offset $\Delta_i$ of the \ac{pdf} peak at pre-contrast,
(ii) a time offset $\Delta_t$ depending of the contrast agent arrival, and
(iii) a change of scale $\alpha_i$ related to the signal enhancement.
Therefore, our normalization method should attenuate all these variations and be performed globally across the different time sequence rather than for each independent sequence.

@@

Therefore, these offsets are estimated by finding the shortest-path to cross the graph using Dijkstra's algorithm.
The entry and exiting nodes are set to be the bin with the maximum probability for the first \ac{dce}-\ac{mri} serie and the bin corresponding to the median value for the last \ac{dce}-\ac{mri} serie, respectively.
To ensure a robust estimation of these offsets, the process of finding the shortest-path is iteratively repeated by shifting the data and updating the heatmap as well as the graph $\mathcal{G}$.
The procedure is stopped once that the offset found does not change.
In general, this process is not repeated more than 3 iterations.
Figure~\ref{fig:estimator} illustrates the final estimation of the offsets (i.e., red landmark) found for each \ac{dce}-\ac{mri} serie.
Therefore, each intensity offset is subtracted for each \ac{dce}-\ac{mri}.

@@

\begin{equation}
  f(t) = \sqrt{ \left( \frac{\sum_{n=0}^{N} x(t)_{n}^2}{N}  \right) },
  \label{eq:rmsd}
\end{equation}

@@

\begin{equation}
  \argmin_{\alpha, \tau} = \sum_{t=0}^{N} \left[ T\left(\alpha, \tau, f(t)\right) - \mu(t) \right]^{2} ,
  \label{eq:cost}
\end{equation}

@@

Illustration of the correction applied to each \ac{rmse} patient is shown in Fig.\,\ref{fig:rmseal}.
Once all these parameters have been inferred, the data are shifted as well as scale.

@@

\noindent where $\Conv$ is the convolution operator; $C_t(t)$ and $C_p(t)$ is the concentration of contrast agent in the tissue and in the plasma, respectively; $K_{trans}$, $k_{ep}$, and $v_p$ are the volume transfer constant, the diffusion rate constant, and the plasma volume fraction, respectively.

@@

    All the above methods are rather complex and thus we propose a method which is based on the following simple assumptions:
    (i) all possible \ac{aif} signal candidates should have a similar shape,
    (ii) an high enhancement, and
    (iii) the arteries should be almost round and within a size range.
    Therefore, each slice is clustered into regions using K-means clustering with $k=6$.
    The cluster with the highest enhancement\textemdash i.e. corresponding to the 90\textsuperscript{th} percentile of the maximum of each dynamic signal\textemdash contain the arteries and is selected.
    Finally, regions with an eccentricity smaller than $0.5$ and an area in the range of $[100, 400]$ voxels are kept.
    Additionally, to remove voxels contaminated by partial volume effect, only the $10\%$ most enhanced voxels of the possible candidates are kept as proposed by~\citep{schabel2008uncertainty} and the average signal is computed.
    A summary of the different segmentation steps is presented in Fig.\,\ref{fig:aif}.
    \item[Conversion of \ac{mri} signal intensity to concentration] To estimate the free parameters of the Tofts model (see Eq.\,\eqref{eq:exttofts}), the concentration $C_t(t)$ and $C_p(t)$ need to be computed from the \ac{mri} signal intensity and the \ac{aif} signal, respectively.
      This conversion is based on the equation of the FLASH sequence\textemdash see~\ref{app:signaltoconc} for details\textemdash and is formulated as in Eq.\,\eqref{eq:conv}:
      \begin{equation}
        c(t) = \frac{1}{TR \cdot r_1} \ln\left( \frac{1 - \cos \alpha \cdot S^{*}\frac{s(t)}{S_0}}{1 - S^{*}\frac{s(t)}{S_0}} \right) - \frac{R_{10}}{r_1} ,
        \label{eq:conv}
      \end{equation}
      \noindent with,
      \begin{equation}
        S^{*} = \frac{1 - \exp(- TR \cdot R_{10})}{1 - \cos \alpha \cdot \exp(- TR \cdot R_{10})} ,
        \label{eq:sstarconv}
      \end{equation}
      \noindent where $s(t)$ is the \ac{mri} signal, $S_0$ is the \ac{mri} signal prior to the injection of the contrast media, $\alpha$ is the flip angle, $TR$ is the \acf{tr}, $R_{10}$ is the pre-contrast tissue relaxation time also equal to $\frac{1}{T_{10}}$, $r_1$ is the relaxitivity coefficient of the contrast agent.

>>>>

In \ac{dce}-\ac{mri}, the intensity \ac{pdf} of prostate gland does not follow a unique type of distribution such as Rician or Gaussian distribution, as shown in Fig.\,\ref{subfig:pathhist}.
Indeed, the inter-patient variations are more complex due to the temporal acquisition.
A better representation to observe these variations is to represent the intensity \ac{pdf} of the prostate gland over time--- requiring to segment the prostate ---using a heatmap representation as shown in Fig.\,\ref{subfig:pathhist}.
Analyzing this heatmap representation across patients (see Fig.\,\ref{subfig:pat2}), the following variations are highlighted:
(i) an intensity offset $\Delta_i$ of the \ac{pdf} peak at pre-contrast,
(ii) a time offset $\Delta_t$ depending of the contrast agent arrival, and
(iii) a change of scale $\alpha_i$ related to the signal enhancement.
Therefore, our normalization method should attenuate all these variations and be performed globally across the different time sequence rather than for each independent sequence.

@@

Therefore, these offsets are estimated by finding the shortest-path to cross the graph using Dijkstra's algorithm.
The entry and exiting nodes are set to be the bin with the maximum probability for the first \ac{dce}-\ac{mri} serie and the bin corresponding to the median value for the last \ac{dce}-\ac{mri} serie, respectively.
To ensure a robust estimation of these offsets, the process of finding the shortest-path is iteratively repeated by shifting the data and updating the heatmap as well as the graph $\mathcal{G}$.
The procedure is stopped once that the offset found does not change.
In general, this process is not repeated more than 3 iterations.
The parameter $\alpha$ is set to $0.9$, empirically.
Figure~\ref{fig:estimator} illustrates the final estimation of the offsets (i.e., red landmark) found for each \ac{dce}-\ac{mri} serie.
Therefore, each intensity offset is subtracted for each \ac{dce}-\ac{mri}.

@@

\begin{equation}
  f(t) = \sqrt{ \left( \frac{\sum_{n=1}^{N} x(t)_{n}^2}{N}  \right) },
  \label{eq:rmsd}
\end{equation}

@@

\begin{equation}
  \argmin_{\alpha, \tau} = \sum_{t=1}^{N} \left[ T\left(\alpha, \tau, f(t)\right) - \mu(t) \right]^{2} ,
  \label{eq:cost}
\end{equation}

@@

Illustration of the correction applied to each \ac{rmse} patient is shown in Fig.\,\ref{fig:rmseal}.
Once all these parameters have been inferred, the data are shifted as well as scaled.

@@

\noindent where $\Conv$ is the convolution operator; $C_t(t)$ and $C_p(t)$ are the concentrations of contrast agent in the tissue and in the plasma, respectively; $K_{trans}$, $k_{ep}$, and $v_p$ are the volume transfer constant, the diffusion rate constant, and the plasma volume fraction, respectively.

@@

    All the above methods are rather complex and thus we propose a method which is based on the following simple assumptions:
    (i) all possible \ac{aif} signal candidates should have a similar shape,
    (ii) a high enhancement, and
    (iii) the arteries should be almost round and within a size range.
    Therefore, each slice is clustered into regions using K-means clustering with $k=6$.
    The cluster with the highest enhancement\textemdash i.e. corresponding to the 90\textsuperscript{th} percentile of the maximum of each dynamic signal\textemdash contain the arteries and is selected.
    Finally, regions with an eccentricity smaller than $0.5$ and an area in the range of $[100, 400]$ voxels are kept.
    Additionally, to remove voxels contaminated by partial volume effect, only the $10\%$ most enhanced voxels of the possible candidates are kept as proposed by~\citep{schabel2008uncertainty} and the average signal is computed.
    A summary of the different segmentation steps is presented in Fig.\,\ref{fig:aif}.
    \item[Conversion of \ac{mri} signal intensity to concentration] To estimate the free parameters of the Tofts model (see Eq.\,\eqref{eq:exttofts}), the concentration $C_t(t)$ and $C_p(t)$ need to be computed from the \ac{mri} signal intensity and the \ac{aif} signal, respectively.
      This conversion is based on the equation of the FLASH sequence\textemdash see~\ref{app:signaltoconc} for details\textemdash and is formulated as in Eq.\,\eqref{eq:conv}:
      \begin{equation}
        c(t) = \frac{1}{TR \cdot r_1} \ln\left( \frac{1 - \cos \alpha \cdot S^{*}\frac{s(t)}{S_0}}{1 - S^{*}\frac{s(t)}{S_0}} \right) - \frac{R_{10}}{r_1} ,
        \label{eq:conv}
      \end{equation}
      \noindent with,
      \begin{equation}
        S^{*} = \frac{1 - \exp(- TR \cdot R_{10})}{1 - \cos \alpha \cdot \exp(- TR \cdot R_{10})} ,
        \label{eq:sstarconv}
      \end{equation}
      \noindent where $s(t)$ is the \ac{mri} signal, $S_0$ is the \ac{mri} signal prior to the injection of the contrast media, $\alpha$ is the flip angle, $TR$ is the \acf{tr}, $R_{10}$ is the pre-contrast tissue relaxation time also equal to $\frac{1}{T_{10}}$, $r_1$ is the relaxitivity coefficient of the contrast agent.
